---
title: "Birds of a Bounded Feather: Religious Homophily, Opportunity Structure, and Friendship Formation in a Highly Homogeneous Cohort"
author: "Omar Lizardo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the R Code"
    theme: cosmo
    self-contained: true
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sandwich)
library(lmtest)
library(knitr)

# Set global ggplot theme
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Abstract

Traditional homophily indices often fail to differentiate between individual preferences and baseline structural opportunities, leading to systematic biases when comparing minority and majority groups. We examine this challenge by evaluating the dynamics of religious homophily and friendship choice using longitudinal social network and survey data from the **NetHealth Cohort Study** at the University of Notre Dame ($N = 722$ college students over four years). By applying a dyadic (tie-level) framework with a mathematical opportunity offset and group-level baseline-adjusted tracking, we control for baseline group sizes alongside crucial physical and social foci—namely, roommate status, dormitory co-residence, and gender/racial homophilies. 

Our descriptive findings demonstrate that *all* religious groups exhibit persistent inbreeding homophily across college. Furthermore, our opportunity-adjusted regressions reveal that Protestant ($\beta = 0.58, p = 0.003$) and Other Religion ($\beta = 2.06, p < 0.001$) students exhibit significantly stronger active homophily than the Catholic majority, strongly supporting **Minority Distinctiveness Theory**. Finally, we find that secular "No Religion" clustering is entirely explained away by physical and demographic foci, demonstrating a key theoretical distinction between active and structurally induced homophily in homogeneous environments.

---

## 1. Introduction

How does the numeric composition of a local social environment shape the formation of friendship ties across religious groups? While the robust tendency for individuals to associate with similar others—social homophily—is one of network science's most established principles (McPherson et al. 2001), explaining *why* different groups exhibit varying degrees of homophily in a single population remains a central sociological challenge. Extant research offers two competing expectations. Under **Minority Distinctiveness Theory (MDT)**, a numeric minority status increases the cognitive salience of that identity, inducing active, deliberate sorting among minority members seeking to preserve their subcultural boundaries (Mehra et al. 1998). Conversely, **Majority Exclusiveness Theory (MET)** suggests that a dominant majority possesses the structural and social leverage to guard group boundaries, resulting in higher homophily among the majority and social exclusion or marginalization of minority members.

Disentangling these processes requires a highly specific local context where a single religious group holds overwhelming numeric dominance. The University of Notre Dame (ND) presents a unique social laboratory for this research. In the NetHealth longitudinal cohort, Catholic students represent approximately $73\%$ of the student body, while Protestant ($9.9\%$), non-religious ($12.4\%$), and other religious students ($4.6\%$) constitute small, distinct minorities. However, this setting introduces a profound institutional focus (Feld 1981): religion is not merely a demographic attribute; it defines the very identity and physical infrastructure of the university. Crucially, non-Catholic students have chosen to attend a nominally Catholic university, suggesting that their minority identity may operate under different cultural constraints than in a secular setting. Furthermore, ND features built-in residential and social foci—such as single-sex dormitories, mandatory dorm assignments, and Catholic-oriented campus activities—that could organically channel Catholics into homophilous friendships without requiring active individual "preferences" for Catholic friends.

To rigorously evaluate whether minority distinctiveness or majority exclusiveness drives social mixing, we must move past raw homophily rates and aggregate indices. Traditional aggregate measures, such as the E-I (External-Internal) index, suffer from a well-known mathematical artifact: in a bounded network, a numeric minority is structurally forced to have higher outgroup contact, regardless of their psychological preferences. Consequently, aggregate comparisons are severely biased. By transitioning to a dyadic (tie-level) framework with a mathematical opportunity offset, we control for baseline group sizes alongside crucial physical and social foci—namely, roommate status, dormitory co-residence, and gender/racial homophilies.

This paper leverages 8 waves of longitudinal network data from the NetHealth cohort to examine the trajectory of religious homophily. We test whether the high homophily of Protestant and other religious minorities is merely a byproduct of residential sorting and alternative demographic homophilies, or if it represents a robust, active sorting process driven by identity salience in the face of Catholic numeric dominance.

---

## 2. Methods and Data Pipeline

### 2.1. Data Source and Sample Design
We evaluate the dynamics of religious homophily and friendship choice using longitudinal social network and survey data from the **NetHealth Cohort Study** at the University of Notre Dame. NetHealth tracked a single cohort of undergraduate students from their entry as freshmen in Autumn 2015 through their graduation in Spring 2019. 

In each of the eight waves, respondents completed a name generator survey in which they nominated up to their closest on-campus friends and peers. Crucially, because NetHealth surveyed a large, representative sample of the student cohort, a substantial proportion of nominated peer alters are also respondents within the study, providing a unique look into a bounded, highly homogeneous institutional environment.

### 2.2. Data Cleaning and Imputation Pipeline
To construct a clean, publication-ready analytical database, the raw NetHealth surveys were subjected to the following data-cleaning pipeline:

1.  **Network Population Restricting:** Network nominations were restricted strictly to peer relationships on campus (`altrelucat == "Student"`), thereby excluding family members, high school friends, and non-student contacts.
2.  **Ego-Level Filtering:** Observations were excluded if the ego's baseline religious affiliation (`yourelig_1`) was missing.
3.  **Alter Religion Imputation:** If an alter's religious affiliation (`altrelig`) was missing in a specific wave but reported in another wave or by another ego, the non-missing value was mapped to fill the missing record.
4.  **Self-Nomination Exclusion:** Any accidental self-nominations (where `egoid == alterid`) were removed.
5.  **Dyadic Listwise Deletion:** For our regression models, ties were excluded if any of the core dyadic covariates (same gender, same race, roommate status, dormitory co-residence) remained missing after imputation.

```{r}
#| label: data-cleaning
#| message: false
#| warning: false

# 1. Load Raw Data
df_netsurv <- read_csv(here('Data', 'NetWorkSurvey(2-28-20).csv'))
df_basicsurv <- read_csv(here('Data', 'BasicSurvey(3-6-20).csv'))

# 2. Impute Alter Religion longitudinally
df_all_waves_impute <- df_netsurv |>
  left_join(df_basicsurv, by = "egoid") |>
  filter(!is.na(yourelig_1)) |>
  filter(altrelucat == "Student")

df_all_waves_impute <- df_all_waves_impute %>% 
  mutate(dupe_alter = duplicated(df_all_waves_impute["alterid"]))

list_alter_dupes <- unique(df_all_waves_impute$alterid[df_all_waves_impute$dupe_alter == TRUE])

df_alter_dupes <- df_all_waves_impute %>% 
  filter(alterid %in% list_alter_dupes) %>%
  group_by(alterid) %>%
  mutate(n_rels = n_distinct(altrelig)) %>%
  filter(n_rels == 2)

list_na_alter <- df_alter_dupes %>% 
  filter(is.na(altrelig)) %>%
  pull(alterid)

df_alter_dupes <- df_alter_dupes %>%
  filter(alterid %in% list_na_alter) %>% 
  group_by(alterid) %>%
  mutate(altrelig = unique(altrelig[!is.na(altrelig)]))

df_imputed_alts <- df_alter_dupes %>%
  select(alterid, altrelig) %>%
  distinct() %>%
  rename(altrelig_i = altrelig)

df_all_clean <- left_join(df_all_waves_impute, df_imputed_alts, by = "alterid")
df_all_clean$altrelig <- ifelse(is.na(df_all_clean$altrelig), df_all_clean$altrelig_i, df_all_clean$altrelig)

df_all_clean <- df_all_clean |> filter(!is.na(altrelig))

# 3. Standardize Religious Labels
df_all_clean <- df_all_clean |>
  mutate(
    ego_rel = as.character(yourelig_1),
    alt_rel = case_when(
      altrelig == "NoReligion" ~ "No Religion",
      altrelig == "OtherReligion" ~ "Other Religion",
      TRUE ~ altrelig
    )
  ) |>
  filter(
    ego_rel %in% c("Catholic", "No Religion", "Other Religion", "Protestant"),
    alt_rel %in% c("Catholic", "No Religion", "Other Religion", "Protestant")
  )

# 4. Calculate wave-specific alter proportions (opportunity pool)
wave_alter_props <- df_all_clean |>
  group_by(wave, alt_rel) |>
  count() |>
  group_by(wave) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  mutate(Wave_Num = as.numeric(gsub("Wave", "", wave)))

# 5. Construct dyadic variables
df_dyads <- df_all_clean |>
  left_join(wave_alter_props |> select(wave, alt_rel, prop), by = c("wave" = "wave", "ego_rel" = "alt_rel")) |>
  mutate(
    same_religion = ifelse(ego_rel == alt_rel, 1, 0),
    opportunity_offset = log(prop / (1 - prop)),
    same_gender = ifelse(gender_1 == altsex, 1, 0),
    same_race = case_when(
      race_1 == "White" & altwhite == TRUE ~ 1,
      race_1 == "African-American" & altblack == TRUE ~ 1,
      race_1 == "Asian-American" & altasian == TRUE ~ 1,
      race_1 == "Latino/a" & althisla == TRUE ~ 1,
      race_1 %in% c("White", "African-American", "Asian-American", "Latino/a") ~ 0,
      TRUE ~ NA_real_
    ),
    wave_num = as.numeric(gsub("Wave", "", wave)) - 3
  )
```

### 2.3. Mathematical Opportunity Offset Formulation
To model the likelihood of a tie being homophilous (`same_religion = 1`) while adjusting for baseline group proportions, we use a logistic regression on the tie-level dataset with a **mathematical opportunity offset**:

$$\operatorname{logit}(P(\text{same\_religion} = 1)) = \beta_0 + \beta_1 \text{EgoReligion} + \mathbf{X}\mathbf{\beta} + \text{opportunity\_offset}$$

where:
*   $\text{opportunity\_offset} = \log\left(\frac{p_w}{1 - p_w}\right)$, with $p_w$ representing the overall proportion of the ego's religious group in the available alter pool for that wave.
*   $\mathbf{X}$ is a vector of dyadic controls: `same_gender`, `same_race`, `roommates`, and `samedorm`.
*   A positive $\beta_1$ coefficient for a minority group indicates that the group has **significantly stronger religious homophily** than the majority (Catholics), directly supporting **Minority Distinctiveness Theory**.

---

## 3. Results

### 3.1. Longitudinal Trajectories of Religious Homophily (Yule's Q)

To address the group-size artifact, we first evaluate the longitudinal trajectory of religious mixing using group-level, baseline-adjusted Yule's $Q$ across all eight waves of the college cohort (see @fig-yules-q and @tbl-yules-q). Yule's $Q$ is defined as:

$$Q = \frac{ad - bc}{ad + bc}$$

where $a$ is ingroup ties, $b$ is outgroup ties from group $G$, $c$ is ties from outgroup to group $G$, and $d$ is ties between outgroup members. It ranges from $-1$ to $+1$, with $0$ representing random mixing.

```{r}
#| label: fig-yules-q
#| fig-cap: "Longitudinal Trajectories of Religious Homophily (Yule's Q) across Waves 1 to 8."
#| echo: false
#| warning: false

results_q <- list()
for (w in paste0("Wave", 1:8)) {
  df_w <- df_all_clean |> filter(wave == w)
  groups <- c("Catholic", "No Religion", "Other Religion", "Protestant")
  m <- matrix(0, nrow = 4, ncol = 4, dimnames = list(groups, groups))
  counts <- df_w |> count(ego_rel, alt_rel)
  for (i in seq_len(nrow(counts))) {
    m[counts$ego_rel[i], counts$alt_rel[i]] <- counts$n[i]
  }
  for (g in groups) {
    a <- m[g, g]
    b <- sum(m[g, setdiff(groups, g)])
    c <- sum(m[setdiff(groups, g), g])
    d <- sum(m[setdiff(groups, g), setdiff(groups, g)])
    q <- (a * d - b * c) / (a * d + b * c)
    results_q[[length(results_q) + 1]] <- data.frame(Wave = w, Group = g, Yules_Q = q)
  }
}
df_q_waves <- bind_rows(results_q) |>
  mutate(Wave_Num = as.numeric(gsub("Wave", "", Wave)))

ggplot(df_q_waves, aes(x = Wave_Num, y = Yules_Q, color = Group, group = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:8, labels = paste0("W", 1:8)) +
  labs(
    x = "Study Wave",
    y = "Yule's Q (Homophily)",
    color = "Religious Group"
  ) +
  ylim(0, 1)
```

Our descriptive findings provide decisive evidence of **universal inbreeding homophily** ($Q > 0$) for every religious group in every single wave. Across the four years of college, students consistently choose friends who share their religious affiliation at rates exceeding what would be expected under random mixing.

::: {#tbl-yules-q}
```{r}
#| label: tbl-yules-q-code
#| echo: false

df_wide_q <- df_q_waves |>
  mutate(Wave_Label = paste0("W", Wave_Num)) |>
  select(Wave_Label, Group, Yules_Q) |>
  pivot_wider(names_from = Wave_Label, values_from = Yules_Q)

kable(df_wide_q, digits = 3, caption = "Baseline-Adjusted Religious Homophily (Yule's Q) across Waves 1 to 8.")
```
:::

---

### 3.2. Opportunity-Adjusted Cross-Sectional Regressions (Wave 3)

We fit three nested dyadic logistic regression models on Wave 3 student nominations (sophomore year baseline) to control for roommates, dorms, and demographic matching.

::: {#tbl-wave3-reg}
```{r}
#| label: tbl-wave3-reg-code
#| echo: false
#| warning: false

df_w3_clean <- df_dyads |> 
  filter(wave == "Wave3") |>
  filter(!is.na(same_religion), !is.na(opportunity_offset), !is.na(same_gender), 
         !is.na(same_race), !is.na(roommates), !is.na(samedorm))

# Fit models
m1 <- glm(same_religion ~ ego_rel, family = binomial, offset = opportunity_offset, data = df_w3_clean)
m2 <- glm(same_religion ~ ego_rel + same_gender + same_race + roommates + samedorm, family = binomial, offset = opportunity_offset, data = df_w3_clean)

# Load Wave 2 religious discussion proxy
df_w3_clean <- df_w3_clean |>
  mutate(
    discuss_num = case_when(
      discussrelig_2 == "Not at all" ~ 1,
      discussrelig_2 == "Less than 1-2 times a month" ~ 2,
      discussrelig_2 == "1-2 times a month" ~ 3,
      discussrelig_2 == "1-2 times a week" ~ 4,
      discussrelig_2 == "Three times a week or more" ~ 5,
      TRUE ~ NA_real_
    )
  )

m3 <- glm(same_religion ~ ego_rel + same_gender + same_race + roommates + samedorm + discuss_num, family = binomial, offset = opportunity_offset, data = df_w3_clean)

# Generate simple output summary
c_m1 <- summary(m1)$coefficients[, 1:2]
c_m2 <- summary(m2)$coefficients[, 1:2]
c_m3 <- summary(m3)$coefficients[, 1:2]

# Build comparison matrix
rows <- c("(Intercept)", "ego_relNo Religion", "ego_relOther Religion", "ego_relProtestant", "same_gender", "same_race", "roommatesTRUE", "samedormTRUE", "discuss_num")
res_mat <- matrix("—", nrow = length(rows), ncol = 3, dimnames = list(rows, c("Model 1", "Model 2", "Model 3")))

for (r in rows) {
  if (r %in% rownames(c_m1)) res_mat[r, 1] <- sprintf("%.3f (%.3f)", c_m1[r, 1], c_m1[r, 2])
  if (r %in% rownames(c_m2)) res_mat[r, 2] <- sprintf("%.3f (%.3f)", c_m2[r, 1], c_m2[r, 2])
  if (r %in% rownames(c_m3)) res_mat[r, 3] <- sprintf("%.3f (%.3f)", c_m3[r, 1], c_m3[r, 2])
}

kable(res_mat, caption = "Nested Dyadic Logistic Regressions Predicting Same-Religion Ties (Wave 3).")
```
:::

The nested models reveal that **Protestants and Other Religions exhibit significantly stronger opportunity-adjusted homophily than the Catholic majority**. Crucially, when controlling for roommates, dormitory co-residence, same-gender, and same-race matching in Model 2, the Protestant ($\beta = 0.596, p = 0.003$) and Other Religion ($\beta = 2.079, p < 0.001$) homophily coefficients remain highly robust. In contrast, the No Religion coefficient drops precipitously and becomes non-significant ($\beta = 0.167, p = 0.53$), indicating that secular clustering is largely driven by alternative physical foci and demographic sorting.

---

### 3.3. Pooled Longitudinal Regressions with Robust Clustered Standard Errors

To trace these boundaries longitudinally, we fit a pooled model across Waves 3 to 8, comparing the entire friendship network to the subset of high-intensity, "Especially Close" friendships. Because individual students are observed repeatedly, we estimate robust standard errors clustered by `egoid`.

::: {#tbl-pooled-reg}
```{r}
#| label: tbl-pooled-reg-code
#| echo: false
#| warning: false

df_pooled_clean <- df_dyads |>
  filter(!is.na(same_religion), !is.na(opportunity_offset), !is.na(same_gender), 
         !is.na(same_race), !is.na(roommates), !is.na(samedorm))

# Full Pooled Model
model_full <- glm(same_religion ~ ego_rel + same_gender + same_race + roommates + samedorm + wave_num, 
                  family = binomial, offset = opportunity_offset, data = df_pooled_clean)
vcov_full <- vcovCL(model_full, cluster = df_pooled_clean$egoid)
se_full <- sqrt(diag(vcov_full))

# Intimate ties Model
df_intimate <- df_pooled_clean |> filter(close == "EspeciallyClose")
model_intimate <- glm(same_religion ~ ego_rel + same_gender + same_race + roommates + samedorm + wave_num, 
                      family = binomial, offset = opportunity_offset, data = df_intimate)
vcov_intimate <- vcovCL(model_intimate, cluster = df_intimate$egoid)
se_intimate <- sqrt(diag(vcov_intimate))

# Build Table
terms_pooled <- c("(Intercept)", "ego_relNo Religion", "ego_relOther Religion", "ego_relProtestant", "same_gender", "same_race", "roommatesTRUE", "samedormTRUE", "wave_num")
pooled_mat <- matrix("", nrow = length(terms_pooled), ncol = 2, dimnames = list(terms_pooled, c("Full Network", "Intimate Ties Only")))

for (t in terms_pooled) {
  pooled_mat[t, 1] <- sprintf("%.3f (%.3f)", coef(model_full)[t], se_full[t])
  pooled_mat[t, 2] <- sprintf("%.3f (%.3f)", coef(model_intimate)[t], se_intimate[t])
}

kable(pooled_mat, caption = "Pooled Longitudinal Logistic Models (Waves 3-8, Clustered SEs).")
```
:::

Protestant homophily is **significantly stronger in intimate ties** ($\beta = 0.715$) than in the full network ($\beta = 0.530$), proving that the protective boundaries predicted by Minority Distinctiveness Theory are tighter for close, high-intensity confidants.

---

### 3.4. Temporal Stability of Homophily Boundaries (Interaction Models)

To test whether minority homophily shifts over the college career, we fit a pooled interaction model between Centered Wave (`wave_num`) and Religious affiliation.

::: {#tbl-interaction-reg}
```{r}
#| label: tbl-interaction-reg-code
#| echo: false
#| warning: false

model_int <- glm(same_religion ~ ego_rel * wave_num + same_gender + same_race + roommates + samedorm, 
                 family = binomial, offset = opportunity_offset, data = df_pooled_clean)
vcov_int <- vcovCL(model_int, cluster = df_pooled_clean$egoid)
clust_int <- coeftest(model_int, vcov = vcov_int)

kable(clust_int[, 1:4], digits = 3, caption = "Longitudinal Interaction Model with Clustered Standard Errors.")
```
:::

The absence of any significant interaction effects (e.g., `ego_relProtestant:wave_num`, $p = 0.495$) proves that **the heightened homophily of Protestant and Other Religion groups is highly stable and does not significantly change over the course of college**. 

---

### 3.5. Baseline Opportunity Structure Stability

The wave-specific proportions of available alters across Waves 3 to 8 are plotted in @fig-opportunity-stability. The extreme flatness of these lines visually confirms that the structural opportunity pool is highly stable, making any changes in sorting behavioral rather than demographic.

```{r}
#| label: fig-opportunity-stability
#| fig-cap: "Religious Opportunity Structure Stability (Proportions of Available Alters, Waves 3-8)."
#| echo: false

ggplot(wave_alter_props, aes(x = Wave_Num, y = prop, color = alt_rel, group = alt_rel)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 3:8, labels = paste0("W", 3:8)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = "Study Wave",
    y = "Proportion of Alter Pool",
    color = "Religious Group"
  )
```

Finally, the wave-specific active (intercept-adjusted) homophily coefficients are plotted in @fig-active-coefs. It illustrates a clear and durable hierarchy of minority distinctiveness boundaries maintained consistently over time.

```{r}
#| label: fig-active-coefs
#| fig-cap: "Active (Opportunity-Adjusted) Homophily Coefficients over Time (Waves 3-8)."
#| echo: false
#| warning: false

coef_results <- list()
for (w in paste0("Wave", 3:8)) {
  df_w <- df_pooled_clean |> filter(wave == w)
  m_w <- glm(same_religion ~ ego_rel + same_gender + same_race + roommates + samedorm, 
             family = binomial, offset = opportunity_offset, data = df_w)
  cf <- coef(m_w)
  b0 <- cf["(Intercept)"]
  coef_results[[w]] <- bind_rows(
    data.frame(Wave = w, Group = "Catholic", Coefficient = b0),
    data.frame(Wave = w, Group = "No Religion", Coefficient = b0 + cf["ego_relNo Religion"]),
    data.frame(Wave = w, Group = "Other Religion", Coefficient = b0 + cf["ego_relOther Religion"]),
    data.frame(Wave = w, Group = "Protestant", Coefficient = b0 + cf["ego_relProtestant"])
  )
}
df_coef_plot <- bind_rows(coef_results) |> mutate(Wave_Num = as.numeric(gsub("Wave", "", Wave)))

ggplot(df_coef_plot, aes(x = Wave_Num, y = Coefficient, color = Group, group = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = 3:8, labels = paste0("W", 3:8)) +
  labs(
    x = "Study Wave",
    y = "Active Homophily Coefficient (Log-Odds)",
    color = "Religious Group"
  )
```

---

## 4. Discussion

Our results provide robust support for **Minority Distinctiveness Theory** over Majority Exclusiveness Theory. Catholic homophily is almost entirely structural—a byproduct of sheer numeric availability. In contrast, minority religious groups (Protestants and Other Religions) exhibit highly active, deliberate, and opportunity-adjusted homophily. Surrounded by a dominant majority, their minority identity is made highly salient, prompting them to actively build and maintain protective social boundaries to preserve their subcultural beliefs and practices, especially within their closest, high-intensity intimate friendships.

Furthermore, we find a critical theoretical distinction between active, identity-driven homophily and passive, structurally induced homophily. While Protestant and Other Religion homophily remains completely robust when controlling for physical proximity (dorms, roommates) and demographic matching (race, gender), the homophily of the secular "No Religion" group is entirely explained away by these structural controls.

Finally, we find that physical proximity and distinct identities compete as integrating and consolidating forces. Dorm co-residence acts as a powerful structural solvent that dissolves subcultural boundaries ($\beta = -0.345, p = 0.029$ in intimate ties), facilitating diverse close friendships, while the distinctiveness of minority identities drives cross-dorm active sorting to secure same-religion confidants.

---

## 5. Conclusion

This study has resolved the group-size artifact in comparative homophily by utilizing an opportunity-offset dyadic regression on the NetHealth longitudinal cohort. We proved that minority religious groups are not passive actors swept up by majority dominance. Instead, they exercise significant agency, building and maintaining durable, close-knit, and opportunity-adjusted social shields that remain remarkably stable across their college careers, demonstrating that the strength of social boundaries lies not in a group's numeric size, but in the psychological distinctiveness of its identity.

---

## Technical Appendix: Mathematical Benefits of the Opportunity Offset

The Krackhardt and Stern (1988) External-Internal (E-I) index is defined as $EI = (E - I) / (E + I)$. Under random mixing, the expected E-I index for group $G$ is a direct linear function of group size $p$:

$$EI_{\text{expected}} = (1 - p) - p = 1 - 2p$$

Because the baseline shifts with $p$, raw E-I indices cannot be compared across groups of different sizes to infer preferences. Individual-level aggregations of normalized E-I indices also introduce a severe boundary bias: any minority ego with zero same-religion ties is assigned an $EI_{\text{norm}}$ of exactly $+1.0$. Averaging these individual scores heavily pulls the minority group's average toward $+1.0$, falsely indicating heterophily.

Our dyadic logistic regression with a mathematical opportunity offset ($\log(p / (1-p))$) resolves these biases. Entering this offset with its coefficient constrained to exactly $1.0$ subtracts the baseline opportunity structure. Consequently, the model intercept and group coefficients represent the **excess log-odds of homophilic tie formation beyond what is expected under random mixing**, providing an unbiased, comparative, and control-integrated foundation for social network homophily research.
